---
title: "Conflict data"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
date: "2023-10-05"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  eval = TRUE,
  results = "asis"
)
```

## Intro

This data takes a look at ACLED conflict data for Somalia. The main focus is to compare 2022 to 2023. Additionally, we wan't to make visuals w/ respect to the [2023 HNO report](https://reliefweb.int/report/somalia/somalia-humanitarian-needs-overview-2023-february-2023) table (p. 55) which identifes priority areas and risk scores for both `Inter-Clan` and  `Non international armed conflict - Al Shabaab` violence.

```{r}
library(tidyverse)
library(gt)
library(rhdx)
library(janitor)
library(ggfx)
library(acled.api)
library(gghdx)
gghdx()

gdf_adm1 <- search_datasets(query = "cod-ab-som") %>%
  pluck(1) %>%
  get_resource(2) %>%
  # get_resource_layers() %>%
  read_resource(layer = "som_admbnda_adm1_ocha_20230308") %>%
  clean_names() %>%
  select(matches("^adm\\d_"))

df_acled_long <- acled.api(
  country = "Somalia",
  start.date = "2010-01-01",
  access.key = Sys.getenv("ACLED_ACCESS_KEY"),
  email.address = Sys.getenv("ACLED_USER_EMAIL"),
  end.date = Sys.Date()
)

# filter to just violent events
df_acled_violent <- df_acled_long %>%
  filter(event_type %in% c(
    "Battles",
    "Riots",
    "Violence against civilians"
  )) %>%
  tibble()
```


From 2023 HNO we have the following classifications

```{r}
violence_class_labels <- c(
  "Non
international
armed
conflict - Al
Shabaab",
  "Inter-clan
violence"
)

al_shabab <- c(
  "Galmudug",
  "Hirshabelle",
  "Jubaland",
  "South West State",
  "Puntland"
)

inter_clan <- c(
  "Galmudug",
  "Hiraan",
  "Lower Shabelle",
  "Middle Shabelle",
  "Sool regions"
)

conflict_priority_zones <- data.frame(
  al_shabab,
  inter_clan
) %>%
  arrange(al_shabab, inter_clan)

colnames(conflict_priority_zones) <- violence_class_labels


conflict_priority_zones %>%
  gt()
```


## 2022 vs 2023 violence

### Absolute Numbers (bar chart)

```{r}
df_acled_violent %>%
  filter(year %in% c(2022, 2023)) %>%
  group_by(year, admin1) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x))
  ) %>%
  ggplot(
    aes(x = reorder(admin1, fatalities), y = fatalities, fill = as_factor(year))
  ) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Somalia ACLED Data",
    subtitle = "Fatalities per admin 1 2022 vs 2023"
  ) +
  theme(
    legend.title = element_blank(),
    axis.title.y = element_blank()
  )
```

### Differnce in years (bar chart)

```{r}
#  cut-off 2022 data at the same month-day as 2023 data ends
df_fatalities_year_recent_norm <- df_acled_violent %>%
  filter(year %in% c(2022, 2023)) %>%
  filter(
    !(year == 2022 & event_date >= "2022-09-29")
  ) %>%
  group_by(year, admin1) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x))
  ) %>%
  select(-interaction) %>%
  pivot_wider(names_from = year, values_from = fatalities) %>%
  mutate(
    diff = `2023` - `2022`,
    diff_class = ifelse(diff < 0, "neg", "pos")
  )



df_fatalities_year_recent_norm %>%
  ggplot(
    aes(x = admin1, y = diff, fill = diff_class)
  ) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = seq(-600, 700, 50)) +
  coord_flip() +
  scale_fill_manual(values = c(hdx_hex("sapphire-hdx"), hdx_hex("tomato-hdx"))) +
  labs(
    title = "Somalia ACLED Data",
    subtitle = "Increase/Decrease Fatalities 2022-2023 (up to sep 29 each year)"
  ) +
  theme(
    legend.title = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 90),
    axis.title.x = element_blank()
  )
```

## HNO 2023 - Groupings for Geographical Scope on Violence Scores

```{r}
# HNO report mentioned does not group always by admin 1 sometimes admin 1's are grouped together to create regions or sometimes
# "states" nonetheless neither of these groupings are CODs so we make them here ourselves
# The groupings used in the report are different depending on whether they are talking about inter-clan or al shabab related violence
# We will still visualize the results at the admin 1 level, but these will be useful to show

gdf_adm1_class <- gdf_adm1 %>%
  mutate(
    hno_intl_conflict = case_when(
      # International & Inter-clan
      adm1_en %in% c("Mudug", "Galgaduud") ~ "Galmadug",

      # International
      # Hirshabelle
      adm1_en %in% c("Middle Shabelle", "Hiraan") ~ "Hirshabelle",
      # Jubaland
      adm1_en %in% c("Gedo", "Lower Juba", "Middle Juba") ~ "Jubaland",
      # Puntland
      adm1_en %in% c("Bari") ~ "Puntland",
      # South WestState
      adm1_en %in% c("Lower Shabelle", "Bay", "Bakool") ~ "South West State",
      .default = NA
    ),
    hno_interclan_conflict = case_when(
      # International & Inter-clan
      adm1_en %in% c("Mudug", "Galgaduud") ~ "Galmadug",

      # Hiran
      adm1_en %in% c("Hiraan") ~ "Hiraan",
      # Middle Shabelle
      adm1_en %in% c("Middle Shabelle") ~ "Middle Shabelle",
      adm1_en %in% c("Lower Shabelle") ~ "Lower Shabelle",

      # Sool
      adm1_en %in% c("Sool") ~ "Sool Regions",
      .default = NA
    ),
  )
```

## Non-International Armed Forces (Al-Shabab) Grouping

```{r}      
gdf_adm1_class %>%
  ggplot() +
  geom_sf(
    aes(
      fill = hno_intl_conflict
    )
  ) +
  scale_fill_discrete(
    name = "Non-Intl\nArmed Conflict",
    na.translate = F
  ) +
  labs(
    title = "Somalia - Regions highlighted for Non-Intl Armed Conflict (Al Shabab)",
    subtitle = "Groupings in legend from 2023 HNO"
  )
```


## Interclan violence grouping
```{r}
gdf_adm1_class %>%
  ggplot() +
  geom_sf(
    aes(
      fill = hno_interclan_conflict
    )
  ) +
  scale_fill_discrete(
    name = "Interclan\nViolence", na.translate = F
  ) +
  labs(
    title = "Somalia - Regions highlighted for inter-clan violence",
    subtitle = "Groupings in legend from 2023 HNO"
  )
```

## Map - Increase/Decrease 2022-2023 in Violence

### Map with Non-Intl Armed Conflict (Al Shabab) boundaries
```{r}
df_fatalities_year_recent_norm %>%
  filter(!admin1 %in% gdf_adm1_class$adm1_en)


gdf_adm1_acled <- gdf_adm1_class %>%
  left_join(df_fatalities_year_recent_norm, by = c("adm1_en" = "admin1"))

gdf_intl_groups_dissolved <- gdf_adm1_acled %>%
  filter(!is.na(hno_intl_conflict)) %>%
  group_by(
    hno_intl_conflict
  ) %>%
  summarise()


ggplot() +
  geom_sf(
    data = gdf_adm1_acled,
    aes(fill = diff),
    alpha = 1
  ) +
  geom_sf(
    data = gdf_adm1_acled %>%
      filter(!is.na(hno_intl_conflict)),
    aes(fill = diff),
    alpha = 1
  ) +
  with_shadow(
    geom_sf(
      data = gdf_intl_groups_dissolved,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 1
    ),
    sigma = 3, x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#709AE1", high = "#FD7446") +
  with_shadow(
    geom_sf_text(
      data = gdf_intl_groups_dissolved,
      aes(label = str_wrap(hno_intl_conflict, width = 8)),
      color = "black"
    ),
    x_offset = 0.2,
    y_offset = 0.2,
    colour = "white", sigma = 2
  ) +
  labs(
    title = "Somalia - Change in Non-Intl Armed Conflict Related Fatalities 2022-2023",
    subtitle = "Regions highlighted in black are groupings used in 2023 HNO"
  )
```

### Map with Interclan Boundaries

```{r}
df_fatalities_year_recent_norm %>%
  filter(!admin1 %in% gdf_adm1_class$adm1_en)



gdf_interclan_groups_dissolved <- gdf_adm1_acled %>%
  filter(!is.na(hno_interclan_conflict)) %>%
  group_by(
    hno_interclan_conflict
  ) %>%
  summarise()

ggplot() +
  geom_sf(
    data = gdf_adm1_acled,
    aes(fill = diff),
    alpha = 1
  ) +
  geom_sf(
    data = gdf_adm1_acled %>%
      filter(!is.na(hno_interclan_conflict)),
    aes(fill = diff),
    alpha = 1
  ) +
  with_shadow(
    geom_sf(
      data = gdf_interclan_groups_dissolved,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 1
    ),
    sigma = 3, x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#709AE1", high = "#FD7446") +
  with_shadow(
    geom_sf_text(
      data = gdf_interclan_groups_dissolved,
      aes(label = str_wrap(hno_interclan_conflict, width = 8)),
      color = "black"
    ),
    x_offset = 0.2,
    y_offset = 0.2,
    colour = "white", sigma = 2
  ) +
  labs(
    title = "Somalia - Change in Interclan Related Fatalities 2022-2023",
    subtitle = "Regions highlighted in black are groupings used in 2023 HNO"
  )
```



