---
title: "Floodscan Explore"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 3
        code_folding: hide
date: "2023-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE, 
  eval = TRUE,
  results = "hide",
  out.width= "100%"
)
```

# Intro

This is the start of an analysis looking at the utility of FloodScan for this research. This analysis is largely **incomplete** and most likely out of scope of project goals so will put on pause.

```{r libsData}
library(tidyverse)
library(sf)
library(tidync)
library(rhdx)
library(exactextractr)
library(janitor)
library(terra)
library(glue)
library(exactextractr)
library(here)
library(gghdx)
# these are for appendix
library(readxl) # read in prmn
library(gganimate)
library(magick)
library(ggridges)
source(here("R/utils_floodscan.R"))
gghdx()

som_adm0_levels <- c(
  "som_admbnda_adm0_ocha_20230308",
  "som_admbnda_adm1_ocha_20230308",
  "som_admbnda_adm2_ocha_20230308"
)

som_adm <- som_adm0_levels %>%
  map(
    ~ search_datasets("Somalia - Subnational Administrative Boundaries") %>%
      pluck(1) %>%
      get_resource(2) %>%
      read_resource(layer = .x) %>%
      clean_names() %>%
      select(matches("^adm\\d_"))
  ) %>%
  set_names(c("adm0", "adm1", "adm2"))

fp_fs <-   file.path(
    Sys.getenv("AA_DATA_DIR"),
    "private",
    "raw",
    "glb",
    "floodscan",
    "floodscan_flooded_fraction_africa_19980112-20221231_p00",
    "aer_sfed_area_300s_19980112_20221231_v05r01.nc"
  )



tnc_fs <- tidync(fp_fs)
tnc_fs_filt <- fs_filter_bounds(fs_obj = tnc_fs,geometry = som_adm$adm0)
r_fs <- fs_to_raster(fs_obj = tnc_fs_filt,band = "SFED_AREA")

dir_wp <-  file.path(
  Sys.getenv("AA_DATA_DIR"),
  "public",
  "raw",
  "som",
  "worldpop"
)
fp_wp <-  file.path(
  dir_wp,
  "som_ppp_2020_1km_Aggregated_UNadj.tif" 
)
r_wp <-  rast(fp_wp)

```

```{r dfElNino}
# this is just for qual visuals
# El Nino year - Just from Wikipedia for a first glance
df_el_nino_years <- tibble(
  date = 
    c("1982-1983",
      "1997-1998",
      "2002-2003",
      "2004-2005",
      "2006-2007",
      "2009-2010",
      "2014-2016",
      "2018-2019",
      "2023-2024"
    )
) %>% 
  separate(
    date ,into = c("start_date","end_date"),sep = "-"
  ) 

# fomat data.frame for ggplot
df_el_nino_years <- df_el_nino_years %>% 
  mutate(
    start_date= glue("{start_date}-01-01") %>% as_date(),
    end_date = glue("{end_date}-12-31") %>% as_date()
  )
```

# Population Exposed to Flooding (FloodScan + World Pop)

Population exposed to flooding is derived by combining FloodScan and World Pop in the following process:

1. Calculate maximum flood scan fraction per a. month and, b. year at the pixel level (max composites)

```{r calcFSMaxComposites}
# This chunk calculates maximum floodscan fraction per month and year

# make a lookup table to be used for raster manipulations
fs_mos<- floor_date(as_date(names(r_fs)),"month")

fs_lookup <- tibble(
  fs_name = as_date(names(r_fs))
) %>% 
  mutate(
    fs_mo = floor_date(fs_name,"month"),
    fs_yr = floor_date(fs_name,"year"),
    fs_seas = paste0(case_when(
      month(fs_mo) %in% c(3,4,5) ~ "MAM",
      month(fs_mo) %in% c(10,11,12) ~ "OND",
      .default ="other"
      
    ),"_",year(fs_yr))
  ) 

# calculate max floodscan fraction per month (pixel level)
lr_monthly_max <- unique(fs_lookup$fs_mo) %>% 
  map(
    \(mo_tmp){
      rname_temp<- fs_lookup %>% 
        filter(fs_mo==mo_tmp) %>% 
        pull(fs_name)
      r_fs_tmp <- r_fs[[names(r_fs) %in% rname_temp]]
      r_max <- max(r_fs_tmp)
      r_max %>% 
        set.names(mo_tmp)
      return(r_max)
    }
  )
# convert list of max rasters back to raster stack
r_fs_monthly_max <-  rast(lr_monthly_max)

# do the same, but this time per year
lr_yearly_max <- unique(fs_lookup$fs_yr) %>% 
  map(
    \(yr_tmp){
      rname_temp<- fs_lookup %>% 
        filter(fs_yr==yr_tmp) %>% 
        pull(fs_name)
      r_fs_tmp <- r_fs[[names(r_fs) %in% rname_temp]]
      r_max <- max(r_fs_tmp)
      r_max %>% 
        set.names(yr_tmp)
      return(r_max)
    }
  )

r_fs_yearly_max <-  rast(lr_yearly_max)
```

2. Align and resample FloodScan max composites to World Pop raster

```{r preProcessFloodScan}
#' Process max composites so they can be combined w/ world pop raster
#' Set extents & resample

# set extents to be equal
r_fs_monthly_max_crop <- crop(r_fs_monthly_max, r_wp)
r_fs_yearly_max_crop <- crop(r_fs_yearly_max, r_wp)
ext(r_fs_monthly_max_crop) <- ext(r_wp)
ext(r_fs_yearly_max_crop) <- ext(r_wp)

system.time(
r_fs_monthly_resampled <- terra::resample(x = r_fs_monthly_max_crop,r_wp)  
)

system.time(
r_fs_yearly_resampled <- terra::resample(x = r_fs_yearly_max_crop,r_wp)  
)

```

3. Multiply resampled FloodScan max composites by World Pop population estimate raster. From this we have a new rasters representing the population exposed per month/year.
```{r multFloodScanWorldPop}
#' Multiply resampled floodscan and world pop data
r_fs_mo_wp<- r_fs_monthly_resampled * r_wp
r_fs_yr_wp<- r_fs_yearly_resampled * r_wp
```

## Appendix

Can we create a baseline?
```{r}
baseline_years <-  1998:2016
fs_lookup_seas <- fs_lookup %>% 
  filter(
    !str_detect(fs_seas,"^other_")
  )

split_rstack_baseline <- function(r, baseline_years){
  ret <- list()
  ret$baseline <- r[[year(names(r)) %in% baseline_years]]
  ret$not_baseline <- r[[!year(names(r)) %in% baseline_years]]
  return(ret)
}
# do the same, but this time per year
lr_seas_max <- unique(fs_lookup_seas$fs_seas) %>% 
  map(
    \(seas_tmp){
      rname_temp<- fs_lookup_seas %>% 
        filter(fs_seas==seas_tmp) %>% 
        pull(fs_name)
      r_fs_tmp <- r_fs[[names(r_fs) %in% rname_temp]]
      r_max <- max(r_fs_tmp)
      r_max %>% 
        set.names(rname_temp[1])
      return(r_max)
    }
  )
r_seas_max <- rast(lr_seas_max)
# set extents to be equal
r_fs_seas_max_crop <- crop(r_seas_max, r_wp)
ext(r_fs_seas_max_crop) <- ext(r_wp)


system.time(
r_fs_seas_max_resampled <- terra::resample(x = r_fs_seas_max_crop,r_wp)  
)

r_fs_seas_mam_max <- r_fs_seas_max_resampled[[month(names(r_fs_seas_max_resampled)) %in% 3]]
r_fs_seas_ond_max <- r_fs_seas_max_resampled[[month(names(r_fs_seas_max_resampled)) %in% 10]]

r_exposure_seas_mam_max <- r_fs_seas_mam_max * r_wp
r_exposure_seas_ond_max <- r_fs_seas_ond_max * r_wp

#rstack 

# mam
rs_exposure_seas_mam <- split_rstack_baseline(r_exposure_seas_mam_max,baseline_years = baseline_years)
r_pop_exposed_mam_baseline_mean <- mean(rs_exposure_seas_mam$baseline) 
r_pop_exposed_mam_baseline_mean %>% 
  set.names("MAM_baseline")

rs_exposure_seas_ond <- split_rstack_baseline(r_exposure_seas_ond_max,baseline_years = baseline_years)
r_pop_exposed_ond_baseline_mean <- mean(rs_exposure_seas_ond$baseline)
r_pop_exposed_ond_baseline_mean %>% 
  set.names("OND_baseline")
  
r_pop_exposed_seas_baselines<- rast(
  list(r_pop_exposed_mam_baseline_mean,
  r_pop_exposed_ond_baseline_mean)
     )

writeRaster(r_pop_exposed_seas_baselines,"somalia_fs_pop_baselines.tif")
```

```{r}
# lets get some stats
df_adm2_historical_exposure_stats_wide <- exact_extract(x = r_exposure_seas_mam_max,
                                y = som_adm$adm2, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode","adm2_en","adm2_pcode"),
                                force_df = TRUE
                                ) 
df_adm2_historical_exposure_stats <- df_adm2_historical_exposure_stats_wide %>% 
  pivot_longer(cols = starts_with("sum"),
               names_to = "sum_date",
               values_to = "pop_exposed") %>% 
  separate(col = sum_date, "\\.",into = c("stat","date")) %>% 
  mutate(
    date= as_date(date)
  )


df_adm2_historical_exposure_stats %>% 
  group_by(
    across(starts_with("adm"))
  ) %>% 
    summarise(
      across(pop_exposed,
             list(mean = mean, min= min, max= max))
      )

df_adm2_historical_exposure_stats %>% 
  ggplot(
    aes(x= pop_exposed,
        y= adm2_en)
  )+
  geom_density_ridges()+
  facet_wrap(~adm1_en,scales="free_y")
  


df_adm1_stats <- df_adm2_stats %>% 
  group_by(adm1_en,adm1_pcode) %>% 
  summarise(
    across(starts_with("sum"),sum),
    .groups="drop"
  ) 
  

```


```{r}
# from ERP
# datapasta::tribble_paste()
df_adm2_erp<- tibble::tribble(
             ~State,           ~adm1_en,       ~adm2_en, ~`2023.HRP.Target`, ~Flood.Affected,
          "Banadir",         "Banadir",       "Banadir",          "902,094",            9300,
         "Galmudug",       "Galgaduud",       "Cadaado",           "31,047",             307,
         "Galmudug",       "Galgaduud", "Dhuusamarreeb",           "63,579",           32212,
         "Galmudug",           "Mudug",     "Gaalkacyo",          "167,697",           17661,
         "Galmudug",           "Mudug",      "Galdogob",           "30,239",              96,
         "Galmudug",           "Mudug",         "Hobyo",           "26,636",           33001,
     "Hirshabellle",          "Hiraan",   "Belet Weyne",           "88,685",          199212,
     "Hirshabellle",          "Hiraan",    "Bulo Burto",            "7,893",           23086,
     "Hirshabellle",          "Hiraan",     "Jalalaqsi",            "7,333",           21618,
     "Hirshabellle", "Middle Shabelle",        "Jowhar",           "99,154",          287572,
     "Hirshabellle", "Middle Shabelle",        "Balcad",           "38,995",          106028,
         "Jubaland",            "Gedo",   "Belet Xaawo",           "26,000",           12301,
         "Jubaland",     "Middle Juba",       "Bu'aale",           "12,635",           73057,
         "Jubaland",      "Lower Juba",       "Afmadow",           "32,776",           12875,
         "Jubaland",            "Gedo",        "Doolow",           "87,622",           97317,
         "Jubaland",            "Gedo",    "Baardheere",           "54,017",           42306,
         "Jubaland",            "Gedo",   "Garbahaarey",           "14,161",           36694,
         "Jubaland",      "Lower Juba",       "Jamaame",           "38,962",          140167,
         "Jubaland",     "Middle Juba",         "Jilib",           "15,311",           49067,
         "Jubaland",      "Lower Juba",      "Kismaayo",           "88,606",           22068,
         "Jubaland",            "Gedo",          "Luuq",           "25,225",          103580,
         "Jubaland",     "Middle Juba",        "Saakow",           "13,108",           17170,
          "Putland",            "Bari",       "Bossaso",          "157,114",             180,
          "Putland",          "Nugaal",      "Burtinle",           "19,093",           13020,
          "Putland",          "Nugaal",       "Garoowe",           "66,218",            5100,
          "Putland",            "Bari",       "Qandala",                "-",             110,
          "Putland",            "Bari",        "Qardho",           "30,849",           62101,
          "Putland",            "Bari",   "Bandarbeyla",            "4,546",             212,
       "South West",             "Bay",      "Baydhaba",          "362,560",           45301,
       "South West",  "Lower Shabelle",       "Afgooye",           "86,438",          101914,
       "South West",          "Bakool",    "Ceel Barde",           "16,029",             720,
       "South West",  "Lower Shabelle",  "Kurtunwaarey",            "9,609",           20742,
       "South West",  "Lower Shabelle",         "Marka",           "55,196",            5201,
       "South West",  "Lower Shabelle",     "Qoryooley",           "12,873",           19431,
       "South West",  "Lower Shabelle",    "Wanla Weyn",           "36,475",           18247,
       "South West",          "Bakool",         "Xudur",           "21,028",             412
     )
```


```{r eval=FALSE}
# quick output for QGIS mapping
fp_swalim <-   file.path(
    Sys.getenv("AA_DATA_DIR"),
    "public",
    "raw",
    "som",
    "swalim"
    
    
  )
gdb_swalim_flood <-  file.path(fp_swalim,"FloodProneAreas_ElninoFloodForecast_HQ100Floo")
gdf_fl_swalim<- st_read(gdb_swalim_flood,"FloodProneAreas_ElninoFloodForecast_HQ100Floo")
gdf_fl_swalim<- st_make_valid(gdf_fl_swalim)
st_intersection(gdf_fl_swalim,
                som_adm$adm2) 
gdf_fl_swalim_adm2<- st_intersection(
  som_adm$adm2,
  gdf_fl_swalim
                ) 

fao_swalim_flooded_adm2 <- gdf_fl_swalim_adm2$adm2_pcode %>% 
  unique()

df_adm2_stats <- exact_extract(x = r_pop_exposed_seas_baselines,
                                y = som_adm$adm2, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode","adm2_en","adm2_pcode")
                                )

df_adm1_stats <- df_adm2_stats %>% 
  group_by(adm1_en,adm1_pcode) %>% 
  summarise(
    across(starts_with("sum"),sum),
    .groups="drop"
  ) 

df_adm1_erp <- df_adm2_erp %>% 
  group_by(adm1_en ) %>% 
  mutate(
    hrp_target_2023= parse_number(str_replace_all(`2023.HRP.Target`,",",""))
  ) %>% 
  summarise(
    across(where(is.numeric),sum)
  ) %>% 
    arrange(desc(Flood.Affected)) %>% 
  mutate(
    flood_rank = row_number()
  )
df_adm1_erp %>% 
  left_join(
    df_adm1_stats
  ) %>%
  clean_names() %>% 
  select(
    adm1_en,
    erp_affected= flood_affected,
    fs_mam_baseline_effected = sum_mam_baseline,
    fs_ond_baseline_effected = sum_ond_baseline,
  ) %>% 
  mutate(
    erp_rank = dense_rank(desc(erp_affected)),
    fs_mam_rank = dense_rank(desc(fs_mam_baseline_effected)),
    fs_ond_rank = dense_rank(desc(fs_ond_baseline_effected))
  )
df_adm1_stats %>% 
  left_join(
    df_adm1_erp
  ) %>%
  clean_names() %>% 
  select(
    adm1_en,
    erp_affected= flood_affected,
    fs_mam_baseline_effected = sum_mam_baseline,
    fs_ond_baseline_effected = sum_ond_baseline,
  ) %>% 
  arrange(desc(fs_mam_baseline_effected)) %>% 
  mutate(
    erp_rank = dense_rank(desc(erp_affected)),
    fs_mam_rank = dense_rank(desc(fs_mam_baseline_effected)),
    fs_ond_rank = dense_rank(desc(fs_ond_baseline_effected))
  )




som_adm$adm2 %>% 
  mutate(
    area= st_area(.)
  ) %>% 
  left_join(
   df_adm2_stats %>% 
     mutate(
       lgl_swalim_100yr = adm2_pcode %in%fao_swalim_flooded_adm2
     )
  ) %>% 
  mutate(
    mam_exposed_pct = sum.MAM_baseline/area,
    ond_exposed_pct = sum.OND_baseline/area
  ) %>% 
  st_write(
    "som_flood.gpkg",
    layer = "adm2_flood_pop_mam_ond",
    delete_layer = T
  )




```




```{r}
library(tidyterra)
library(gghdx)
gghdx()

adm1_labs <- som_adm$adm1 %>% 
  mutate(
       CENTROID = map(geometry, st_centroid),
    COORDS = map(CENTROID, st_coordinates),
    COORDS_X = map_dbl(COORDS, 1),
    COORDS_Y = map_dbl(COORDS, 2)
  )
adm2_labs <- som_adm$adm2 %>% 
  mutate(
       CENTROID = map(geometry, st_centroid),
    COORDS = map(CENTROID, st_coordinates),
    COORDS_X = map_dbl(COORDS, 1),
    COORDS_Y = map_dbl(COORDS, 2)
  )

values(r_pop_exposed_seas_baselines[[2]] ) %>% 
  data.frame() %>% 
  ggplot(aes(
    x= OND_baseline
  ))+
  geom_histogram()+
  scale_x_continuous(trans = scales::pseudo_log_trans())

r_plot_pop_exposed_seas_baselines <- deepcopy(r_pop_exposed_seas_baselines)
r_plot_pop_exposed_seas_baselines[r_plot_pop_exposed_seas_baselines<10] <- NA
RColorBrewer::brewer.pal("YlGnBu",n=7)
ggplot()+
  geom_spatraster(data=r_plot_pop_exposed_seas_baselines)+
  # scale_fill_gradient(
  #   trans= scales::pseudo_log_trans(),
    # na.value = NA,
    # low = "white",
    # high="red",
  #   breaks = c(0,1,10,500),
  #   name="Population\nExposed"
  #   ) +
  scale_fill_stepsn(
    trans= scales::pseudo_log_trans(),
    colours=c(RColorBrewer::brewer.pal("YlGnBu",n=3)),
    breaks= c(10,25,50,100),
    na.value=NA
  )+
  # scale_fill_binned(
  #       na.value = NA,
  #   low = "white",
  #   mid="orange",
  #   high="darkred",
  #   breaks= c(0,1,5,10,20,50,100,150,200),
  #   name= "Population Exposed"
  # )+
    facet_wrap(~lyr)+
  geom_sf(data=som_adm$adm2, 
          fill= NA,
          color=alpha("lightgrey",0.2),
          lwd=0.5
          )+
  geom_sf(data=som_adm$adm1, 
          fill= NA,
          color="grey",
          lwd=0.6)+
  geom_sf(data=som_adm$adm0, 
          fill= NA,
          color="black")+
   geom_text_repel(
    data= adm1_labs,
    mapping = aes(
      x = COORDS_X,
      y = COORDS_Y,
      label = adm1_en
    ),
     size= 6,
    color="black"
  )+
  labs(
    title = "Average Flood Exposure Per Year (1998-2016)",
    subtitle = "Somalia",
      caption = "Cumulative flood exposure calculated per year and multiplied by world pop 2020 UN adjusted raster to calculate total exposure per year. The average exposure was then calculated between 1998-2016."
  )+
  theme_hdx()+
  theme(
    plot.background = element_rect(color="white",fill="white"),
    plot.title = element_text(size=24),
    plot.subtitle = element_text(size=20),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.title=element_blank(),
    axis.line=element_blank(),
    plot.caption=element_text(hjust= 0, size= 10),
    legend.key.size = unit(0.1, "in"),
  )
ggsave(filename = "teset.png",width = 5,height = 6,units = "in",dpi = 300)
```


```{r}

con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = "global_gdb",
                      # user      = rstudioapi::askForPassword("Database user"),
                      # password      = rstudioapi::askForPassword("Database password"),
                      port     = 5432)

DBI::dbListTables(con)
gdf_h2r<- st_read(con, "som_h2r_feb_drought_indicators" )


# Baidoa zoom in 
gdf_adm1_bay <- som_adm$adm1 %>% 
  filter(adm1_en=="Bay") 
  
bbox_adm1_bay <- gdf_adm1_bay %>% 
  st_bbox()
  
gdf_bbox_adm1_bay <- bbox_adm1_bay %>% 
  st_as_sfc()

gdf_ch_adm1_bay <- gdf_adm1_bay %>% 
  st_convex_hull()

gdf_h2r_crop = st_crop(gdf_h2r,gdf_bbox_adm1_bay)

ggplot()+
  geom_spatraster(data=r_plot_pop_exposed_seas_baselines)+
  scale_fill_stepsn(
    trans= scales::pseudo_log_trans(),
    colours=c(RColorBrewer::brewer.pal("YlGnBu",n=3)),
    breaks= c(10,25,50,100),
    na.value=NA
  )+
    facet_wrap(~lyr)+
  geom_sf(data=som_adm$adm2, 
          fill= NA,
          color=alpha("lightgrey",0.2),
          lwd=0.5
          )+
  geom_sf(data=som_adm$adm1, 
          fill= NA,
          color="grey",
          lwd=0.6)+
  geom_sf(data=som_adm$adm0, 
          fill= NA,
          color="black")+
  geom_sf(
    data= gdf_h2r_crop,
    color= alpha("lightgrey",0.5),
    size= 0.3, alpha= 0.5, shape=4
  )+
   geom_text_repel(
    data= st_crop(adm2_labs,gdf_ch_adm1_bay),
    mapping = aes(
      x = COORDS_X,
      y = COORDS_Y,
      label = adm2_en
    ),
     size= 4,
    color="black"
  )+
   geom_text_repel(
    data= st_crop(adm1_labs,gdf_bbox_adm1_bay),
    mapping = aes(
      x = COORDS_X,
      y = COORDS_Y,
      label = toupper(adm1_en)
    ),
     size= 8,
    color="grey"
  )+
    coord_sf(
    xlim=c(adm1_bay[["xmin"]],adm1_bay[["xmax"]]),
    ylim=c(adm1_bay[["ymin"]],adm1_bay[["ymax"]]),
    )+
  labs(
    title = "Average Flood Exposure Per Year (1998-2016)",
    subtitle = "Somalia",
      caption = "Cumulative flood exposure calculated per year and multiplied by world pop 2020 UN adjusted raster to calculate total exposure per year. The average exposure was then calculated between 1998-2016."
  )+
  theme_hdx()+
  theme(
    plot.background = element_rect(color="grey",fill="white"),
    panel.background = element_rect(color="grey", fill=NA),
    plot.title = element_text(size=24),
    plot.subtitle = element_text(size=20),
    panel.grid = element_blank(),
    # panel.border = ele(),
    axis.text = element_blank(),
    axis.title=element_blank(),
    axis.line=element_blank(),
    plot.caption=element_text(hjust= 0, size= 10),
    legend.key.size = unit(0.1, "in"),
  )
ggsave(filename = "bay.png",width = 4,height = 4,units = "in",dpi = 300)
```


```{R}



p_map_ond_baseline <- ggplot()+
  geom_spatraster(data=r_pop_exposed_ond_baseline_mean)+
  scale_fill_gradient(
    trans= scales::pseudo_log_trans(),
    na.value = NA,
    low = "white",
    high="red",breaks = c(0,5,10,20,50,100,150,200),
    name="Population\nExposed"
    ) +
  geom_sf(data=som_adm$adm2, 
          fill= NA,
          color=alpha("lightgrey",0.2),
          lwd=0.5
          )+
  geom_sf(data=som_adm$adm1, 
          fill= NA,
          color="grey",
          lwd=0.6)+
  geom_sf(data=som_adm$adm0, 
          fill= NA,
          color="black")+
   geom_text_repel(
    data= adm1_labs,
    mapping = aes(
      x = COORDS_X,
      y = COORDS_Y,
      label = adm1_en
    ),
     size= 6
  )+
  labs(
    title = "Average Flood Exposure Per Year (1998-2016)",
    subtitle = "Somalia"
  )+
  theme_hdx()+
  theme(
    plot.background = element_rect(color="white",fill="white"),
    plot.title = element_text(size=20),
    plot.subtitle = element_text(size=18),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.title=element_blank(),
    axis.line=element_blank()
  )
```



```{r}




#rstack 
rs_pop_exposed <- split_rstack_baseline(r_fs_yr_wp,baseline_years = baseline_years)
r_pop_exposed_baseline_mean <- mean(rs_pop_exposed$baseline)

rs_flood_frac <-  split_rstack_baseline(r = r_fs_yearly_max_crop,baseline_years = baseline_years)
r_flood_frac_baseline <-  mean(rs_flood_frac$baseline)

```

```{r}
library(tidyterra)
library(gghdx)
gghdx()

adm1_labs <- som_adm$adm1 %>% 
  mutate(
       CENTROID = map(geometry, st_centroid),
    COORDS = map(CENTROID, st_coordinates),
    COORDS_X = map_dbl(COORDS, 1),
    COORDS_Y = map_dbl(COORDS, 2)
  )


ggplot()+
  geom_spatraster(data=r_pop_exposed_baseline_mean)+
  scale_fill_gradient(
    trans= scales::pseudo_log_trans(),
    na.value = NA,
    low = "white",
    high="red",breaks = c(0,5,10,20,50,100,150,200),
    name="Population\nExposed"
    ) +
  geom_sf(data=som_adm$adm2, 
          fill= NA,
          color=alpha("lightgrey",0.2),
          lwd=0.5
          )+
  geom_sf(data=som_adm$adm1, 
          fill= NA,
          color="grey",
          lwd=0.6)+
  geom_sf(data=som_adm$adm0, 
          fill= NA,
          color="black")+
   geom_text_repel(
    data= adm1_labs,
    mapping = aes(
      x = COORDS_X,
      y = COORDS_Y,
      label = adm1_en
    ),
     size= 6
  )+
  labs(
    title = "Average Flood Exposure Per Year (1998-2016)",
    subtitle = "Somalia",
    caption = "Cumulative flood exposure calculated per year and multiplied by world pop 2020 UN adjusted raster to calculate total exposure per year. The average exposure was then calculated between 1998-2016"
  )+
  theme_hdx()+
  theme(
    plot.background = element_rect(color="white",fill="white"),
    plot.title = element_text(size=20),
    plot.subtitle = element_text(size=18),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.title=element_blank(),
    axis.line=element_blank()
  )
ggsave(filename = "teset.png",width = 5,height = 6,units = "in",dpi = 300)
```

```{r})
```


4. From the new rasters we can simply run zonal sums per admin zone of interest. Here we do it mainly by admin 1.

```{r calcZonalStats}

# Monthly Pop Exposed Per Admin 1
df_adm1_mo_pop_exposed<- exact_extract(x = r_fs_mo_wp,
                                y = som_adm$adm1, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode")
                                ) %>% 
  pivot_longer(-matches("adm")) %>%
  separate(name,
           into = c("stat", "date"),
           sep = "\\.") %>% 
  mutate(
    date= as_date(date)
  )

# Monthly Pop Exposed per Admin 2
df_adm2_mo_pop_exposed<- exact_extract(x = r_fs_mo_wp,
                                y = som_adm$adm2, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode","adm2_en","adm2_pcode")
                                ) %>% 
  pivot_longer(-matches("adm")) %>%
  separate(name,
           into = c("stat", "date"),
           sep = "\\.")%>% 
  mutate(
    date= as_date(date)
  )


# Yearly Pop Exposed per Admin 1
df_adm1_yr_pop_exposed <- exact_extract(x = r_fs_yr_wp,
                                y = som_adm$adm1, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode")
                                ) %>% 
  pivot_longer(-matches("adm")) %>%
  separate(name,
           into = c("stat", "date"),
           sep = "\\.") %>% 
  mutate(
    date= as_date(date)
  )



# World Pop Adin 1 Stats (Population per admin 1)
df_adm1_wp<- exact_extract(x = r_wp,
                                y = som_adm$adm1, 
                                fun ="sum",
                                append_cols= c("adm1_en","adm1_pcode")
                                ) %>% 
  pivot_longer(-matches("adm")) %>%
  separate(name,
           into = c("stat", "date"),
           sep = "\\.") %>% 
  mutate(date=as_date(date))

```

## Plot Population Exposed per Month
```{r plot-monthlyExposedAdm1, results= "asis"}
# Population expose
df_adm1_mo_pop_exposed %>% 
  ggplot(
    aes(x= date, y = value, group=adm1_en)
    )+
  geom_line()+
  scale_y_continuous(labels = scales::label_comma())+
  scale_x_date(
    date_breaks = "1 year",
    date_labels= "%y"
  )+
  labs(title = "Monthly population exposed to flooding",
       y= "Population exposed",
       caption= "Population: World Pop\nFlood: FloodScan")+
  facet_wrap(~adm1_en)+
  theme(
    axis.text.x = element_text(angle=90, size=7),
    axis.title.x =  element_blank(),
    plot.caption = element_text(hjust=0 , size=7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
  )
```

## Plot Population Exposed per year

```{r plot-yearlyExposedAdm1, results="asis"}
df_adm1_yr_pop_exposed %>% 
  ggplot(aes(x= date, y = value, group=adm1_en))+
  geom_line()+
  scale_y_continuous(labels = scales::label_comma())+
  scale_x_date(
    date_breaks = "1 year",
    date_labels= "%y"
  )+
  labs(title = "Yearly population exposed to flooding",
       y= "Population exposed",
       caption= "Population: World Pop\nFlood: FloodScan")+
  facet_wrap(~adm1_en)+
  theme(
    axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
  )
```

## Flood Exposure Anomalies

From the admin level population exposed rasters we can easily calculate two types of anomalies:

1. Monthly difference from mean expsure
2. Monthly Z-score (difference from mean/ standard deviation)

```{r calcFloodExposureAnom}
# calc anomaly
df_adm1_mo_pop_exposed_anom <- df_adm1_mo_pop_exposed %>% 
  mutate(
    mo = month(date, label=T)
  ) %>% 
  group_by(adm1_en, adm1_pcode, mo) %>% 
  mutate(
    mean = mean(value),
    stdev = sd(value),
    diff_mean = value-mean,
    z_score = diff_mean/stdev
  ) %>% 
  ungroup()
```

### Plot difference from mean exposure

```{r plot-monthlyDiffAnomaly,results="asis"}
# difference from mean anomaly
df_adm1_mo_pop_exposed_anom %>% 
  ggplot(aes(x= date, y = diff_mean, group=adm1_en))+
  geom_line()+
  scale_y_continuous(labels = scales::label_comma())+
  scale_x_date(
    date_breaks = "1 year",
    date_labels= "%y"
  )+
    labs(title = "Monthly Flood Exposure Anomaly",
       y= "Population exposed Anomaly (difference from average)",
       caption= "Population: World Pop\nFlood: FloodScan"
       )+
  facet_wrap(~adm1_en)+
  theme(
    axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
  )
```

### Plot Exposure Anomaly (Z-score)

```{r plot-MonthlyZscore, results="asis"}

df_adm1_mo_pop_exposed_anom %>% 
  ggplot()+
  geom_line(
    aes(x= date,
        y = z_score,
        group=adm1_en)
  )+
  scale_y_continuous(
    labels = scales::label_comma()
  )+
  labs(title = "Monthly Flood Exposure Anomaly",
       y= "Population exposed Anomaly (Z-score)",
       caption= "Population: World Pop\nFlood: FloodScan"
       )+
  facet_wrap(~adm1_en)+
  geom_rect(
    data= df_el_nino_years,
    aes(
      xmin = start_date,
      xmax = end_date,
      ymin = -Inf,
      ymax = Inf
    ),
    alpha = 0.2,
    fill = hdx_hex("tomato-hdx")
  )+
    scale_x_date(
    date_breaks = "1 year",
    date_labels= "%y",
    limits = c(min(df_adm1_mo_pop_exposed_anom$date),as_date("2025-06-01")),
    expand = c(0,0)
  )+
  theme(
   axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
  )
```

# Appendix

## FloodScan WorldPop Exposure Vs PRMN

```{r readPRMN}

dir_som_pub_raw <- file.path(
  Sys.getenv("AA_DATA_DIR"),
  "public",
  "raw",
  "som")

fp_prmn <- file.path(dir_som_pub_raw,
          "prmn",
          "UNHCR-PRMN-Displacement-Dataset.xlsx"
          )
df_prmn <- read_xlsx(fp_prmn) %>% 
  clean_names()

df_prmn <-  df_prmn %>% 
  filter(reason=="Flood") %>% 
  mutate(
    date= dmy(month_end),
    yr_mo = floor_date(date, "month"),
    .before= everything()
  )

df_prmn_adm1_mo <- df_prmn %>% 
  group_by(
    adm1_en = current_arrival_region,
    date= yr_mo
    ) %>% 
  summarise(
   number_of_individuals= sum(number_of_individuals,na.rm=T) ,
   .groups="drop"
  )
```


```{r joinFloodScanPRMN}
df_fs_prmn <- df_adm1_mo_pop_exposed_anom %>% 
  left_join(
    df_prmn_adm1_mo,
    by = c("date"="date",
           "adm1_en"="adm1_en")
  ) 
```

```{r plot-FloodScanVsPrmn, fig.height=9, results="asis"}
df_fs_prmn %>% 
  ggplot(aes(x= value, y= number_of_individuals))+
  geom_point(size=0.7)+
  scale_y_continuous(labels = scales::label_comma())+
  scale_x_continuous(labels = scales::label_comma())+
  labs(
    title = "Floodscan/World Pop vs PRMN Displacement by Admin 1",
    x= "Pop Exposure (FloodScan + World Pop)",
    y= "Number Displaced (PRMN)",
    caption= "Population: World Pop\nFlood: FloodScan\nDisplacement: UNHCR PRMN"
  )+
  facet_wrap(~adm1_en, scale="free")+
  theme(
    axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
  )
```


```{r, eval =F}
df_fs_prmn_seas <- df_fs_prmn %>% 
  mutate(
    seas = case_when(
      month(date) %in% c(3,4,5)~"MAM",
      month(date)%in% c(10,11,12)~"OND"
    )
  ) %>% 
  filter(!is.na(seas)) %>% 
  group_by(across(matches("adm")),
                  yr=floor_date(date,"year"),seas) %>% 
  summarise(
    number_of_individuals= sum(number_of_individuals,na.rm=T),
    mean_z= mean(z_score),
    max_z= max(z_score),.groups="drop"
  ) 


df_fs_prmn_seas %>% 
  filter(seas== "MAM") %>% 
  ggplot(aes(x= mean_z, y= number_of_individuals))+
  geom_point()+
  facet_wrap(~adm1_en, scales="free")


df_fs_prmn_seas %>% 
  filter(seas== "OND") %>% 
  ggplot(aes(x= mean_z, y= number_of_individuals))+
  geom_point()+
  facet_wrap(~adm1_en,scales = "free")

# max OND
df_fs_prmn_seas %>% 
  filter(seas== "OND") %>% 
  ggplot(aes(x= max_z, y= number_of_individuals))+
  geom_point()+
  facet_wrap(~adm1_en,scales="free")
```


## Baidoa case study
```{r wrangleBaidoa}
df_prmn_baidoa_fs <- 
  left_join(
    df_adm2_mo_pop_exposed %>% 
      filter(year(date)>=2016,
             adm2_en=="Baydhaba") ,
  df_prmn %>% 
  filter(
   previous_departure_district== "Baidoa"
  ) %>% 
     group_by(date= yr_mo) %>% 
  summarise(
    number_of_individuals= sum(number_of_individuals,na.rm=T)
  ) ,
  by = c("date"="date")
  )
```

### Plot - Pop displaced & Exposed
```{r plot-BaidoaPopExposed,results="asis"}
df_prmn_baidoa_fs  %>% 
  select(
    adm2_en, date, `PRMN Displaced individuals`=number_of_individuals, `FloodScan + WorldPop Displaced` =value
  ) %>% 
  pivot_longer(`PRMN Displaced individuals`:`FloodScan + WorldPop Displaced`,
               names_to = "variable",
               values_to = "value") %>% 
  ggplot(
    aes(x= date, 
        y= value, 
        fill= variable, 
        group= variable, 
        color = variable
    )
    
    )+
  geom_point(alpha=0.8)+
  geom_line(alpha=0.5)+
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")+
    labs(
    title = "Baidoa - PRMN Displacement and FloodScan/WorldPop Exposure",
    y= "Exposed/Displaced",
    caption= "Population: World Pop\nFlood: FloodScan\nDisplacement: UNHCR PRMN"
  )+
  theme(
   axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
   
  )
```

### Plot - Pop displaced & Exposed
```{r plot-BaidoaDualPopExposed,results="asis"}
df_prmn_baidoa_fs  %>% 
  select(
    adm2_en, date, number_of_individuals, value
  ) %>% 
  ggplot(
    aes(x= date, y= number_of_individuals)
    )+
  geom_point()+
  geom_line()+
   geom_vline(
     data = . %>% 
       filter(!is.na(number_of_individuals)),
    aes(xintercept = date), 
    color="red",alpha= 0.3
  )+
  geom_line(
    data= df_prmn_baidoa_fs,
    aes(x= date, y= value/10),
    color = hdx_hex("mint-hdx")
  )+
  scale_y_continuous(
    sec.axis = sec_axis(~.*10, name = "WorldPop Exposure (thousands)")
    )+
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b%y", 
    limits = as_date(c("2016-01-01","2022-12-01")),expand = c(0,0)
    )+
  labs(
    y= "PRMN - Number of Individuals",
    title= "Somalia Baidoa - Displaced vs Flood Exposure",
    subtitle = "Blue dots and vertical red lines indicate PRMN Flood Displacement",
    caption= "Population: World Pop\nFlood: FloodScan\nDisplacement: UNHCR PRMN"
  )+
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
    
  )
```

### Plot - Pop displaced & Exposed (pseudolog)
```{r plot-BaidoaDualPopExposedPseudo,results="asis"}
df_prmn_baidoa_fs  %>% 
  select(
    adm2_en, date, number_of_individuals, value
  ) %>% 
  ggplot(
    aes(x= date, y= number_of_individuals)
    )+
  geom_point()+
  geom_line()+
   geom_vline(
     data = . %>% 
       filter(!is.na(number_of_individuals)),
    aes(xintercept = date), 
    color="red",alpha= 0.3
  )+
  geom_line(
    data= df_prmn_baidoa_fs,
    aes(x= date, y= value/10)
  )+
  scale_y_continuous(
    trans = scales::pseudo_log_trans(),
    breaks = c(05,25,50,seq(100,1000, 200),2000,3000,9000),
    sec.axis = sec_axis(
      ~.*10, 
      name = "WorldPop Flood Exposure (thousands)",
      breaks= c(50,250,500,1000,10000,20000,30000,60000,90000)
      )
    )+
  scale_x_date(date_breaks = "1 month", date_labels = "%b%y")+
  labs(
    y= "PRMN - Number of Individuals",
    title= "Somalia Baidoa - Displaced vs Flood Exposure",
    subtitle = "Blue dots and vertical red lines indicate PRMN Flood Displacement",
    caption= "Population: World Pop\nFlood: FloodScan\nDisplacement: UNHCR PRMN"
  )+
    theme(
    legend.position = "none",
    axis.text.x = element_text(angle=90),
    axis.text.y = element_text(size= 8),
    axis.title.x =  element_blank(),
    axis.title.y =  element_text(size=8),
    plot.caption = element_text(hjust=0, size= 7),
    strip.text = element_text(size=9),
    panel.background = element_rect(color="grey",fill=NA)
    
  )
  
  
```


```{r animationSetup}
gdf_adm1_yr_pop_exposed <- som_adm$adm1 %>% 
  left_join(
    df_adm1_yr_pop_exposed,
    by = c("adm1_en"="adm1_en")
  ) 

# use ggplot2 to create a map using gdf_adm1_fly_yr
# I want to fill by the `value` column and I want to 
# annimate with gganimate by the `date` column


# this produces : Error: arguments have different crs 
# devtools::install_github("thomasp85/transformr")
# animate so that the color of the fill changes value based on date
# go through all the dates
# https://github.com/thomasp85/gganimate/issues/479

# create a ggplot object
p <- gdf_adm1_yr_pop_exposed %>% 
  ggplot() +
  geom_sf(aes(fill=value)) +
  scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = "bottom") +
  # transition_states and make each state last 1 second 
  transition_states(states = date,
                    # state_length = 2,
                    wrap = FALSE) +
  labs(title = "Flooding Exposure - {closest_state}",
       subtitle = "World Pop + Floodscan",
       fill = "Population Exposed"
       )

map_gif <- animate(p, 
        fps = 2,
        width = 800,
        height = 600)


# create ggplot line plot that adds a new year of data every time from df_zonal_fl_pop_yr 
# and animate with gganimate

p_facet_plot <- df_adm1_yr_pop_exposed %>% 
  mutate(
    date= as_date(date)
  ) %>% 
  ggplot(aes(x= date, y = value))+
  geom_line()+
  scale_y_continuous(labels = scales::label_comma())+
  scale_x_date(
    date_breaks = "1 year",
    date_labels= "%y"
  )+
  labs()+
  facet_wrap(~adm1_en)+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.x =  element_blank()
  )+
  transition_reveal(date)

plot_gif <- animate(p_facet_plot, 
        fps = 2,
        width = 800,
        height = 600)



map_mgif <- image_read(map_gif)
plot_mgif <- image_read(plot_gif)
length(map_mgif)
length(plot_mgif)


new_gif <- image_append(c(map_mgif[1], plot_mgif[1]))

for(i in 2:92){
  print(i)
  combined <- image_append(c(map_mgif[i], plot_mgif[i]))
  new_gif <- c(new_gif, combined)
}
```

```{r FunAnimation,results= "asis"}
new_gif
```






