---
title: "Conflict data"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        toc_depth: 4
date: "2023-10-05"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  eval = TRUE,
  results = "asis"
)
```

# Intro

This data takes a look at ACLED conflict data for Somalia. The main focus is to compare 2022 to 2023. Additionally, we wan't to make visuals w/ respect to the [2023 HNO report](https://reliefweb.int/report/somalia/somalia-humanitarian-needs-overview-2023-february-2023) table (p. 55) which identifies priority areas and risk scores for both `Inter-Clan` and  `Non international armed conflict - Al Shabaab` violence.

The visuals produced here are the ones critical to the conflict-related analysis/visuals presented in this [brainstorming slide-deck](https://docs.google.com/presentation/d/1LuMEILKzSp8Ujp355fItPXC1skFoOalAIhPpJPHWglc/edit#slide=id.g2456d9f415a_0_102)

```{r}
library(tidyverse)
library(gt)
library(rhdx)
library(janitor)
library(ggfx)
library(acled.api)
library(gghdx)
library(sf)
library(patchwork)
gghdx()

gdf_adm1 <- search_datasets(query = "cod-ab-som") %>%
  pluck(1) %>%
  get_resource(2) %>%
  # get_resource_layers() %>%
  read_resource(layer = "som_admbnda_adm1_ocha_20230308") %>%
  clean_names() %>%
  select(matches("^adm\\d_"))

df_acled_long <- acled.api(
  country = "Somalia",
  start.date = "2010-01-01",
  access.key = Sys.getenv("ACLED_ACCESS_KEY"),
  email.address = Sys.getenv("ACLED_USER_EMAIL"),
  end.date = Sys.Date(),
  all.variables=T
)

# filter to just violent events
df_acled_violent <- df_acled_long %>%
  filter(event_type %in% c(
    "Battles",
    "Riots",
    "Violence against civilians"
  )) %>%
  tibble()
```

```{r}
df_acled_violent <- df_acled_violent %>% 
  mutate(
    date= as_date(event_date),
    conflict_class = case_when(
      str_detect(actor1, "Al Shabaab")~"Al Shabaab" ,
      str_detect(actor2, "Al Shabaab")~"Al Shabaab" ,
      str_detect(actor1, "Clan")~"Clan" ,
      str_detect(actor2, "Clan")~"Clan" ,
      str_detect(actor1, "Militia")~"Clan" ,
      str_detect(actor2, "Militia")~"Clan" ,
      .default = "other")
  )


```

# Conflict Bar Charts

## All Fatalities 2022 vs 2023

```{r}
df_acled_violence_gte2022 <- df_acled_violent %>%
  filter(year %in% c(2022, 2023),
         !(year == 2022 & event_date >= "2022-09-29")
         )

df_adm1_acled_gte2022 <- df_acled_violence_gte2022 %>%
  group_by(year, admin1) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x))
  ) 

df_adm1_acled_gte2022 %>%
  ggplot(
    aes(x = reorder(admin1, fatalities), y = fatalities, fill = as_factor(year))
  ) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Somalia ACLED Data",
    subtitle = "Fatalities per admin 1 - 2022 vs 2023"
  ) +
  theme(
    legend.title = element_blank(),
    axis.title.y = element_blank()
  )
```

## Fatalities by Conflict Type - 2023 vs 2023
```{r}
df_adm1_conflict_gte2022 <- df_acled_violence_gte2022 %>%
  filter(
    conflict_class!="other",
  ) %>%
  mutate(
    conflict_class_label = if_else(conflict_class == "Al Shabaab", "Al Shabaab Related", "Clan Related")
  ) %>% 
  group_by(year, admin1,conflict_class_label) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x)),
    .groups="drop"
  )

df_adm1_conflict_gte2022 %>%
  ggplot(
    aes(x = reorder(admin1, fatalities), y = fatalities, fill = as_factor(year))
  ) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  facet_wrap(~conflict_class_label)+
  labs(
    title = "Somalia ACLED Data - By Confict Type",
    subtitle = "Fatalities per admin 1 2022 vs 2023",
    caption = "Classification of conflict type:\n1. If Al-Shabaab was listed as actor 1 or 2 it was event is considered 'Al Shabaab Related'\n2. If Al-Shabaab was not listed and either 'Clan' or 'Militia' was listed as actor 1 or 2 the event was classified as 'Clan Related'\n3. All other events types not considered"
  ) +
  theme(
    legend.title = element_blank(),
    axis.title.y = element_blank(),
    plot.caption = element_text(hjust=0)
  )
```

## Increase/Decrease from 2022 to 2023


ACLED conflict fatalities by type (Shabaab vs Interclan)
```{r}
df_adm1_conflict_acled_compare %>%
  ggplot(
    aes(x = admin1, y = diff, fill = diff_class)
  ) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = seq(-600, 700, 50)) +
  coord_flip() +
  scale_fill_manual(values = c(hdx_hex("sapphire-hdx"), hdx_hex("tomato-hdx"))) +
  labs(
    title = "Somalia ACLED Data",
    subtitle = "Increase/Decrease Fatalities 2022-2023",
    caption = "Note: scales of x-axis differ in 2 plots"
  ) +
  geom_text( aes(
    label=diff,
    y=ifelse(diff<=100, diff+ 50* sign(diff),diff + (.05*diff))
    # y=max(diff)+.05*max(diff)
                 ),
    angle=270)+
  facet_wrap(~conflict_class_label,scales = "free_x")+
  theme(
    legend.title = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 90),
    axis.title.x = element_blank(),
    plot.caption = element_text(hjust=0),
    panel.border = element_rect(color="grey",fill=NA)
  )
  
```

# Maps

From 2023 HNO we have the following classifications

```{r}
violence_class_labels <- c(
  "Non
international
armed
conflict - Al
Shabaab",
  "Inter-clan
violence"
)

al_shabab <- c(
  "Galmudug",
  "Hirshabelle",
  "Jubaland",
  "South West State",
  "Puntland"
)

inter_clan <- c(
  "Galmudug",
  "Hiraan",
  "Lower Shabelle",
  "Middle Shabelle",
  "Sool regions"
)

conflict_priority_zones <- data.frame(
  al_shabab,
  inter_clan
) %>%
  arrange(al_shabab, inter_clan)

colnames(conflict_priority_zones) <- violence_class_labels


conflict_priority_zones %>%
  gt()
```

## HNO 2023 - Groupings for Geographical Scope on Violence Scores

```{r}
# HNO report mentioned does not group always by admin 1 sometimes admin 1's are grouped together to create regions or sometimes
# "states" nonetheless neither of these groupings are CODs so we make them here ourselves
# The groupings used in the report are different depending on whether they are talking about inter-clan or al shabab related violence
# We will still visualize the results at the admin 1 level, but these will be useful to show

gdf_adm1_class <- gdf_adm1 %>%
  mutate(
    hno_intl_conflict = case_when(
      # International & Inter-clan
      adm1_en %in% c("Mudug", "Galgaduud") ~ "Galmadug",

      # International
      # Hirshabelle
      adm1_en %in% c("Middle Shabelle", "Hiraan") ~ "Hirshabelle",
      # Jubaland
      adm1_en %in% c("Gedo", "Lower Juba", "Middle Juba") ~ "Jubaland",
      # Puntland
      adm1_en %in% c("Bari") ~ "Puntland",
      # South WestState
      adm1_en %in% c("Lower Shabelle", "Bay", "Bakool") ~ "South West State",
      .default = NA
    ),
    hno_interclan_conflict = case_when(
      # International & Inter-clan
      adm1_en %in% c("Mudug", "Galgaduud") ~ "Galmadug",

      # Hiran
      adm1_en %in% c("Hiraan") ~ "Hiraan",
      # Middle Shabelle
      adm1_en %in% c("Middle Shabelle") ~ "Middle Shabelle",
      adm1_en %in% c("Lower Shabelle") ~ "Lower Shabelle",

      # Sool
      adm1_en %in% c("Sool") ~ "Sool Regions",
      .default = NA
    ),
  )
```

## Map Admin Grouping: Non-International Armed Forces (Al-Shabab) Grouping

```{r}      
gdf_adm1_class %>%
  ggplot() +
  geom_sf(
    aes(
      fill = hno_intl_conflict
    )
  ) +
  scale_fill_discrete(
    name = "Non-Intl\nArmed Conflict",
    na.translate = F
  ) +
  labs(
    title = "Somalia - Regions highlighted for Non-Intl Armed Conflict (Al Shabab)",
    subtitle = "Groupings in legend from 2023 HNO"
  )
```


## Map Admin Grouping: Interclan violence 
```{r}
gdf_adm1_class %>%
  ggplot() +
  geom_sf(
    aes(
      fill = hno_interclan_conflict
    )
  ) +
  scale_fill_discrete(
    name = "Interclan\nViolence", na.translate = F
  ) +
  labs(
    title = "Somalia - Regions highlighted for inter-clan violence",
    subtitle = "Groupings in legend from 2023 HNO"
  )
```


## Non-Intl Armed Conflict (Al Shabaab) boundaries

### Map - Shabaab

- Below is a map showing the increase decrease in Al-Shabaab related conflict fatalities between 2022 & 2023. Just leaving auto legend as is for now until we decide on more polished visuals.
- Black delineations represent the admin groupings used in the 2023 HNO to define areas of increased Non-Intl Armed conflict (Al Shabaab) related conflict

```{r}

gdf_adm1_acled_shabaab <- gdf_adm1_class %>%
  left_join(df_adm1_conflict_acled_compare %>% 
              filter(
                conflict_class_label =="Al Shabaab Related"
              ), by = c("adm1_en" = "admin1"))


gdf_intl_groups_dissolved <- gdf_adm1_class %>%
  filter(!is.na(hno_intl_conflict)) %>%
  group_by(
    hno_intl_conflict
  ) %>%
  summarise()

ggplot() +
  geom_sf(
    data = gdf_adm1_acled_shabaab,
    aes(fill = diff),
    alpha = 1
  ) +
  with_shadow(
    geom_sf(
      data = gdf_intl_groups_dissolved,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 1
    ),
    sigma = 3, x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#709AE1", high = "#FD7446") +
  with_shadow(
    geom_sf_text(
      data = gdf_intl_groups_dissolved,
      aes(label = str_wrap(hno_intl_conflict, width = 8)),
      color = "black"
    ),
    x_offset = 0.2,
    y_offset = 0.2,
    colour = "white", sigma = 2
  ) +
  labs(
    title = "Somalia - Increase/Decrease in conflict-related fatalities 2022-2023",
    subtitle = "Black lines delineate areas used in the 2023 HNO to define Al-Shabaab related zones",
    caption = "Only fatalities in which Al-Shabaab was involved"
  )+
  theme(
    plot.title = element_text(size=14),
    plot.subtitle = element_text(size=12),
    axis.title = element_blank(),
    plot.caption = element_text(hjust=0)
  )
```

### Trends Al-Shabaab 

Below we can the trend for each admin 1 faceted/grouped by the 2023 HNO admin grouping definitions.

```{r, results='hide'}
latest_date_yday <- yday(
  as_date("2022-09-29")
)

# filter all years so they do not go past Sep 29.
df_acled_violent_norm <- df_acled_violent %>%
  filter(
        yday(event_date)<=latest_date_yday
        ) 
  
# good
df_acled_violent_norm %>% 
  filter(!admin1 %in% gdf_adm1_class$adm1_en) %>% 
  distinct(admin1)

# add HNO groupings
df_acled_violent_norm_hno <- df_acled_violent_norm %>% 
  left_join(
    gdf_adm1_class %>% 
      st_drop_geometry() %>% 
      select(-starts_with("adm0")), 
    by = c("admin1"="adm1_en")
  )


df_adm1_acled_conflict <- df_acled_violent_norm_hno %>%
  mutate(
    hno_intl_conflict = ifelse(is.na(hno_intl_conflict),"Not grouped as Al-Shabaab related",hno_intl_conflict),
    hno_interclan_conflict = ifelse(is.na(hno_interclan_conflict),"Not grouped as Clan related",hno_interclan_conflict)
  ) %>% 
  group_by(year = floor_date(
    as_date(event_date),"year"), hno_intl_conflict,admin1,conflict_class) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x)),
    .groups="drop"
  ) 

#' Because there are so many admin 1 units involved in this grouping it a simple facet_wrap does not look good
#' as there would be too many line colors. I want to split plots by HNO grouping, but preserve the admin 1 information
#' Therefore I can just loop through and use `{patchwork}` rather than ~facet_wrap

hno_shabaab_regions <- df_adm1_acled_conflict$hno_intl_conflict %>% unique()

p_shabaab_trends <- 
  map(set_names(hno_shabaab_regions),
      \(region_name){
    df_adm1_acled_conflict %>% 
      filter(conflict_class=="Al Shabaab",
             hno_intl_conflict==region_name) %>% 
      ggplot(
        aes(x= year, y=fatalities, color= admin1, group=admin1)
      )+
      geom_point()+
      geom_line()+
      scale_x_date(date_breaks ="2 year",
                   date_labels = "%Y")+
      labs(
        title = region_name
      )+
      theme(
        legend.title = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(size=10),
        axis.text.x = element_text(angle=90),
        legend.text = element_text(size=8),legend.key = element_rect(size=3),
        panel.border = element_rect(color="grey", fill = NA)
      )
  }
  )
```

```{r}
p_shabaab_trends$Jubaland+
  p_shabaab_trends$`South West State`+
  p_shabaab_trends$Hirshabelle+
  p_shabaab_trends$Galmadug+
  p_shabaab_trends$Puntland+
  p_shabaab_trends$`Not grouped as Al-Shabaab related`+
  plot_annotation(
    title = "Somalia: Yeary Al-Shabaab Related Conflict Fatalities",
    subtitle = "Split by 2023 HNO Groupings"
    )+
  plot_layout(ncol=3)
```


## Interclan Conflict

### Map Interclan Conflict

- Below is a map showing the increase decrease in Interclan related conflict fatalities between 2022 & 2023. Just leaving auto legend as is for now until we decide on more polished visuals.
- Black delineations represent the admin groupings used in the 2023 HNO to define areas of increased Intercland-violence related conflict risk.
```{r}

gdf_adm1_acled_clan <- gdf_adm1_class %>%
  left_join(df_adm1_conflict_acled_compare %>% 
              filter(
                conflict_class_label =="Clan Related"
              ), by = c("adm1_en" = "admin1"))  



gdf_interclan_groups_dissolved <- gdf_adm1_acled %>%
  filter(!is.na(hno_interclan_conflict)) %>%
  group_by(
    hno_interclan_conflict
  ) %>%
  summarise()

ggplot() +
  geom_sf(
    data = gdf_adm1_acled_clan,
    aes(fill = diff),
    alpha = 1
  ) +
  with_shadow(
    geom_sf(
      data = gdf_interclan_groups_dissolved,
      # aes(fill= diff),
      fill = NA,
      alpha = 1,
      color = "black",
      lwd = 1
    ),
    sigma = 3, x_offset = 0.5,
    y_offset = 0.25
  ) +
  scale_fill_gradient2(low = "#709AE1", high = "#FD7446") +
  with_shadow(
    geom_sf_text(
      data = gdf_interclan_groups_dissolved,
      aes(label = str_wrap(hno_interclan_conflict, width = 8)),
      color = "black"
    ),
    x_offset = 0.2,
    y_offset = 0.2,
    colour = "white", sigma = 2
  ) +
  labs(
    title = "Somalia - Increase/Decrease in Interclan-Confict Related Fatalities between 2022 & 2023",
    subtitle = "Black lines delineate areas used in the 2023 HNO to define zones with higher levels of Interclan violence",
    caption = "Only conflict events in which at least one actor was described as Clan or Militia and no actors were Al-Shabaab\nGrey indicates no Interclan events reported in 2022 or 2023"
  )+
  theme(
    plot.title = element_text(size=13),
    plot.subtitle = element_text(size=10),
    axis.title = element_blank(),
    plot.caption = element_text(hjust=0, size=10)
  )
```

### Trends - Interclan Conflict

Below we can the trend for each admin 1 faceted/grouped by the 2023 HNO admin grouping definitions.
```{r}

df_adm1_acled_conflict_interclan <- df_acled_violent_norm_hno %>%
  mutate(
    hno_intl_conflict = ifelse(is.na(hno_intl_conflict),"Not grouped as Al-Shabaab related",hno_intl_conflict),
    hno_interclan_conflict = ifelse(is.na(hno_interclan_conflict),"Not grouped as Clan related",hno_interclan_conflict)
  ) %>% 
  group_by(year = floor_date(
    as_date(event_date),"year"), hno_interclan_conflict,admin1,conflict_class) %>%
  summarise(
    across(c("interaction", "fatalities"), ~ sum(.x)),
    .groups="drop"
  ) 
```


```{r}
# Simple facet_wrap will work w/ less admin lines to color
df_adm1_acled_conflict_interclan %>% 
  filter(conflict_class=="Clan",
         hno_interclan_conflict!="Not grouped as Clan related") %>% 
  ggplot(
    aes(x= year, y=fatalities, color= admin1, group=admin1)
  )+
  geom_point()+
  geom_line()+
  scale_x_date(date_breaks ="2 year",
               date_labels = "%Y")+
  scale_y_continuous(
    breaks= seq(0,350, 50),
    limits= c(0, 325),
    expand = c(0,0)
  )+
  facet_wrap(~hno_interclan_conflict)+
  labs(
    title = "Somalia: Yeary Interclan Related Conflict Fatalities",
    subtitle = "Split by 2023 HNO Groupings"
  )+
  theme(
    legend.title = element_blank(),
    axis.title = element_blank(),
    
    axis.text.x = element_text(angle=90),
    # legend.text = element_text(size=8),legend.key = element_rect(size=3),
    panel.border = element_rect(color="grey", fill = NA)
  )
```


```{r}

df_adm1_acled_conflict_interclan %>% 
  filter(conflict_class=="Clan",
         hno_interclan_conflict=="Not grouped as Clan related") %>% 
  ggplot(
    aes(x= year, y=fatalities, color= admin1, group=admin1)
  )+
  geom_point()+
  geom_line()+
  scale_x_date(date_breaks ="2 year",
               date_labels = "%Y")+
  scale_y_continuous(
    breaks= seq(0,150, 25),
    limits= c(0, 125),
    expand = c(0,0)
  )+
  labs(
    title = "Somalia: Yeary Interclan Related Conflict Fatalities",
    subtitle = "Small flare ups in clan-related violence in regions not classified as interclan violence zones (HNO 2023)"
  )+
  scale_color_brewer(type = "qual",palette = "Set3")+
  theme(
    legend.title = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(angle=90),
    panel.border = element_rect(color="grey", fill = NA)
  )


```
